Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-10-01 papers/paper1/component_model.tex
Updates through page 10.
presubmission tweaks

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition, and if it is true applying the effect of the output action and all composed input actions using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

Component frameworks that support \emph{recursive encapsulation} allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

@@

\begin{figure}
\center
\includegraphics[width=\columnwidth]{system_model}
\caption{Example event detection system.}
\label{sys_model}
\end{figure}

@@

\paragraph*{Concurrent execution}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming a countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems can be modeled in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.

@@

\paragraph*{Support for dynamic systems}
To address the limitations of the static representation of dynamic systems, ioa++ provides
architectural support for dynamic systems by separating the concerns of \emph{enacting} versus \emph{initiating} 
actions for managing dynamics into distinct \emph{system} and \emph{base} automata respectively.
We define four \emph{system actions} for managing dynamic systems of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
The \emph{create} action allows an automaton to create a child automaton.
The \emph{bind} action allows an automaton to associate an input action with an output action while the \emph{unbind} action allows an automaton to dissociate an input action from an output action.
The \emph{destroy} action allows an automaton to destroy a child automaton and all bindings associated with it (recursively).
\begin{figure}
\center
\includegraphics[width=\columnwidth]{system_action}
\caption{System automaton and actions.}
\label{system_action}
\end{figure}

% A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
% Thus, $S_{t+1}$ need not be equal to $S_{t}$.
% A \emph{system action} is an action that \emph{can} change the system $S$.
% The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions for returning responses to system action requests.
(Parameterized actions are indicated with [].)
The base automaton provides output actions that trigger requests to create, bind, unbind, and destroy automata, along with
input actions to receive the results.
User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class. 
Each user automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.

@@

%% Let $A$ be the set of automata that exist where each element in $A$ is a pair $(p, a)$ where $p$ is the parent of $a$.
%% Let $B$ be the set of bindings that exist where each element in $B$ is a tuple $(w, o_a, o_n, o_p, i_a, i_n, i_p)$ where $w$ is the automaton that owns the binding, $o_a$ is the output automaton, $o_n$ is the name of the output action, $o_p$ is the parameter of the output action, $i_a$ is the input automaton, $i_n$ is the name of the input action, and $i_p$ is the parameter of the input action.
%% Define the predicate $exists (a) = \langle \exists p, q : (p, q) \in A :: q = a\rangle$ that is true when automaton $a$ exists, i.e., has a parent\footnote{The root automaton's parent is the special symbol $\perp$.}.
%% Define the predicate $existsb (a) = \langle \exists a : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: a = w \lor a = o_a \lor a = i_a \rangle$ that is true when automaton $a$ appears somewhere in $B$.
%% We define the create operation as the Hoare triple $\{exists (p) \land \lnot exists (a)\} \quad create(a)_p \quad \{(p,a) \in A\}$ which says that the parent automaton $p$ can create automaton $a$ if $p$ exists and $a$ does not exist.
%% The destroy operation can be defined $\{(p,a) \in A\} \quad destroy(a)_p \quad \{(p,a) \notin A \land \lnot existsb(a)\}$ which says that in order to destroy $a$ and remove all bindings associated with $a$, $p$ must be the parent of $a$.
%% Define the predicate $free (a, n, p) = \langle \forall w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: \lnot (i_a = a \land i_n = n \land i_p = p) \rangle$ that is true when the input action argument is not bound.
%% Define the predicate $involved (o, n, p, i) = \langle \exists w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: o_a = o \land o_n = n \land o_p = p \land i_a = i \rangle$ that is true when automaton $o$'s output action $(n, p, i)$ is bound to some input action in automaton $i$.
%% The bind operation can be defined $\{exists (w) \land exists (o_a) \land exists (i_a) \land free (i_a, i_n, i_p) \land \lnot involved(o_a, o_n, o_p, i_a) \land o_a \neq i_a \} \quad bind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\}$ which says that in order for the binding to succeed the owner $w$ exists, automaton $o_a$ exists, automaton $i_a$ exists, the input action $(i_a, i_n, i_p)$ must not be bound, the output action $(o_n, o_p)$ of automaton $o_a$ cannot be bound to an input action in automaton $i_a$, and the automata $o_a$ and $i_a$ cannot be the same.
%% The precondition for $bind$ is same as the binding rules derived in section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
%% The unbind operation is similar to destroy: $\{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\} \quad unbind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \notin B\}$.

>>>>

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
The scheduler is assumed to be \emph{fair,} meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition, and if it is true applying the effect of the output action and all composed input actions using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

\begin{figure*}
\center
\includegraphics[width=\textwidth]{system_model}
\caption{Example event detection system.}
\label{sys_model}
\end{figure*}
Component frameworks that support \emph{recursive encapsulation} allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

@@

@@

\paragraph*{Concurrent execution}
Recall that composing I/O automata consists of concatenating state variables and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming a countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems can be \emph{modeled} in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.

@@

\paragraph*{Support for dynamic systems}
To address the limitations of the static representation of dynamic systems, ioa++ provides
architectural support for dynamic systems by separating the concerns of \emph{enacting} versus \emph{initiating} 
actions for managing dynamics into distinct \emph{system} and \emph{base} automata respectively.
We define four \emph{system actions} for managing dynamic systems of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
The \emph{create} action allows an automaton to create a child automaton.
The \emph{bind} action allows an automaton to associate an input action with an output action while the \emph{unbind} action allows an automaton to dissociate an input action from an output action.
The \emph{destroy} action allows an automaton to destroy a child automaton and all bindings associated with it (recursively).

\begin{figure*}
\center
\includegraphics[width=\textwidth]{system_action}
\caption{System automaton and actions.}
\label{system_action}
\end{figure*}
% A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
% Thus, $S_{t+1}$ need not be equal to $S_{t}$.
% A \emph{system action} is an action that \emph{can} change the system $S$.
% The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions for returning responses to system action requests.
(Parameterized actions are indicated with [].)
The base automaton provides output actions that trigger requests to create, bind, unbind, and destroy automata, along with
input actions to receive the results.
User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class. 
Each user automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.

@@

%% Let $A$ be the set of automata that exist where each element in $A$ is a pair $(p, a)$ where $p$ is the parent of $a$.
%% Let $B$ be the set of bindings that exist where each element in $B$ is a tuple $(w, o_a, o_n, o_p, i_a, i_n, i_p)$ where $w$ is the automaton that owns the binding, $o_a$ is the output automaton, $o_n$ is the name of the output action, $o_p$ is the parameter of the output action, $i_a$ is the input automaton, $i_n$ is the name of the input action, and $i_p$ is the parameter of the input action.
%% Define the predicate $exists (a) = \langle \exists p, q : (p, q) \in A :: q = a\rangle$ that is true when automaton $a$ exists, i.e., has a parent\footnote{The root automaton's parent is the special symbol $\perp$.}.
%% Define the predicate $existsb (a) = \langle \exists a : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: a = w \lor a = o_a \lor a = i_a \rangle$ that is true when automaton $a$ appears somewhere in $B$.
%% We define the create operation as the Hoare triple $\{exists (p) \land \lnot exists (a)\} \quad create(a)_p \quad \{(p,a) \in A\}$ which says that the parent automaton $p$ can create automaton $a$ if $p$ exists and $a$ does not exist.
%% The destroy operation can be defined $\{(p,a) \in A\} \quad destroy(a)_p \quad \{(p,a) \notin A \land \lnot existsb(a)\}$ which says that in order to destroy $a$ and remove all bindings associated with $a$, $p$ must be the parent of $a$.
%% Define the predicate $free (a, n, p) = \langle \forall w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: \lnot (i_a = a \land i_n = n \land i_p = p) \rangle$ that is true when the input action argument is not bound.
%% Define the predicate $involved (o, n, p, i) = \langle \exists w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: o_a = o \land o_n = n \land o_p = p \land i_a = i \rangle$ that is true when automaton $o$'s output action $(n, p, i)$ is bound to some input action in automaton $i$.
%% The bind operation can be defined $\{exists (w) \land exists (o_a) \land exists (i_a) \land free (i_a, i_n, i_p) \land \lnot involved(o_a, o_n, o_p, i_a) \land o_a \neq i_a \} \quad bind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\}$ which says that in order for the binding to succeed the owner $w$ exists, automaton $o_a$ exists, automaton $i_a$ exists, the input action $(i_a, i_n, i_p)$ must not be bound, the output action $(o_n, o_p)$ of automaton $o_a$ cannot be bound to an input action in automaton $i_a$, and the automata $o_a$ and $i_a$ cannot be the same.
%% The precondition for $bind$ is same as the binding rules derived in Section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
%% The unbind operation is similar to destroy: $\{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\} \quad unbind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \notin B\}$.

### 2011-10-01 papers/paper1/design.tex
presubmission tweaks

The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments .
The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.

@@

Since concurrent access to the dispatcher is allowed, the dispatcher implementation the dispatcher must be thread-safe.
Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

>>>>

The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments.
The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.

@@

Since concurrent access to the dispatcher is allowed, the dispatcher implementation must be thread-safe.
Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.
